\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-31 paper.tex
Problem para

% What is the problem?
The true ARG underlying a modern population-scale dataset
therefore cannot be known in its entirety, and the value
of attempting to describe the intricate details of the recent past
through mathematical models is questionable.
A pragmatic goal that we may set, then, is to infer \emph{aspects}
of the true ARG and to capture uncertainty where it inevitably exists.
This is the approach taken by recent inference
% TODO should we mention SARGE? I don't think it's in the same class, but
% maybe I'm wrong?
methods~\citep{kelleher2019inferring,speidel2019method,zhang2023biobank},
which estimate an approximate version of the true
ARG, informed, but not strictly constrained, by population genetic
models and parsimony.
While these methods have made substantial progress there
are, however, some significant problems in this emerging field.
Our terminology and mathematical formalisms for ARGs,
ultimately derived from the coalescent,
are ill-suited to describing these approximate structures.
% TODO add some citations of examples?
Discussions are forced to contrast a ``full'' ARG, containing
a precise description of every recombination or coalescence event,
with a sequence of disconnected local trees.
This view is depauperate, obscuring many shades of important
and interesting detail between these two extremes.
Although the term ``ARG'' is now often used in a general
sense~\citep[e.g.][]{mathieson2020ancestry,hejase2020summary,
schaefer2021ancestral,harris2023using,zhang2023biobank,fan2023likelihood},
as informally
encompassing the varied outputs from modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
kelleher2019inferring, speidel2019method, baumdicker2021efficient,
zhang2023biobank},
there is no corresponding mathematical description of an ARG data structure.
Classical formalisms, based on the Griffiths approach, are insufficient to
describe these rich details.
Given that ARGs have already been estimated for the 500,000
humans in UKB~\citep{kelleher2019inferring,zhang2023biobank} and
more than a million SARS-CoV-2 genomes~\citep{zhan2023towards},
it is vital that our formal methods and terminology catch up
with this reality.

% % pedantically, ARG is not really a "word"
% The term ``ARG'' is now frequently used in a general
% sense~\citep[e.g.][]{mathieson2020ancestry,hejase2020summary,
% schaefer2021ancestral,harris2023using,zhang2023biobank,fan2023likelihood},
% as a data structure
% encompassing the varied outputs from modern simulation and
% inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
% kelleher2019inferring, speidel2019method, baumdicker2021efficient, zhang2023biobank}.
% This broad usage is a useful generalisation,
% but is not compatible with the available formal
% data structure definitions
% derived from the original Griffiths approach.
% This mismatch between usage and definition has led to significant
% problems, which if not addressed
% may impede progress in applying ARGs to understand the biological world.

% % % Para 3: what is the problem?
% The classical formal definition of an ARG is derived from the
% stochastic process, and expressed in terms of events:
% the common ancestor and recombination events
% that occur in the history of a sample. We show that this approach
% to representing ARGs (which we call the ``event ARG'', or eARG
% encoding) is inherently limiting, insufficient to represent
% the complexities of modern datasets, and incompatible with
% the outputs of modern large-scale simulation and inference
% methods~\citep{kelleher2016efficient,speidel2019method,
% kelleher2019inferring,zhang2023biobank}.
% We address these problems by providing a more flexible formulation
% of an ARG data structure, which we call the ``genome ARG''
% or gARG encoding. We show that it is fully compatible with the
% classical eARG approach, while providing much more flexibility
% in how genetic inheritance can be represented,
% and encompasses the outputs of modern methods.
% Moreover we show that, in contrast with the eARG encoding which
% requires precise estimates of each recombination event,
% the gARG encoding allows us to systematically describe uncertainty
% about the temporal ordering of multiple recombinations.

@@

% We must also distinguish between a specific realisation
% the \emph{actual} ARG underlying a sample of genomes

>>>>

% What is the problem?
The true ARG underlying a modern population-scale dataset
therefore cannot be known in its entirety, and the value
of attempting to describe the intricate details of the recent past
through mathematical models is questionable.
A pragmatic goal that we may set, then, is to infer \emph{aspects}
of the true ARG and to capture uncertainty where it inevitably exists.
This is the approach taken by recent inference
% TODO should we mention SARGE? I don't think it's in the same class, but
% maybe I'm wrong?
methods~\citep{kelleher2019inferring,speidel2019method,zhang2023biobank},
which estimate an approximate version of the true
ARG, informed, but not strictly constrained, by population genetic
models and parsimony.
While these methods have made substantial progress there
are, however, some significant problems in this emerging field.
Our terminology and mathematical formalisms for ARGs,
ultimately derived from the coalescent,
are ill-suited to describing these approximate structures.
% TODO add some citations of examples?
Discussions are forced to contrast a ``full'' ARG, containing
a precise description of every recombination or coalescence event,
with a sequence of disconnected local trees.
This view is depauperate, obscuring many shades of important
and interesting detail between these two extremes.
Although the term ``ARG'' is now often used in a general
sense~\citep[e.g.][]{mathieson2020ancestry,hejase2020summary,
schaefer2021ancestral,harris2023using,zhang2023biobank,fan2023likelihood},
informally
encompassing the varied outputs from modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
kelleher2019inferring, speidel2019method, baumdicker2021efficient,
zhang2023biobank},
there is no corresponding mathematical description of an ARG data structure.
Classical formalisms, based on the Griffiths approach, are insufficient to
describe these rich details.
Given that ARGs have already been estimated for the 500,000
humans in UKB~\citep{kelleher2019inferring,zhang2023biobank} and
more than a million SARS-CoV-2 genomes~\citep{zhan2023towards},
it is vital that our formal methods and terminology catch up
with this reality.

We address these problems here by providing a simple formal description of an
ARG data structure, based on recording specific genomes and their intervals of
genetic inheritance. We call this the ``genome ARG'', or gARG encoding. We
contrast this with the classical formal definition of an ARG, based on
recording common ancestor and recombination events, which we refer to as the
``event ARG'' or eARG encoding. We show that the new gARG encoding it is a
substantial generalisation of the classical eARG approach,  providing much more
flexibility in how genetic inheritance can be represented, and encompassing the
outputs of modern methods. We show that the gARG approach can represent many
different types of approximation, in particular allowing us to systematically
describe uncertainty about the temporal ordering of multiple recombinations. It
is important to note that throughout we are interested in the details of these
competing mathematical formulations and their practical consequences.
% Do we need to say this? Probably not?
We are not questioning the form of the true ARG, or its existance.

@@
